cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Updated

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. These shapes help test how well methods preserve symmetry, curvature, and cluster separation. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:sphere-tb-html)', '\\@ref(tab:sphere-tb-pdf)'))` summarizes these functions.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle." In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = 1/3\cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, $X_3 = ru$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p$.

@@

### Trigonometric

@@

### Trefoil knots

>>>>

Sphere-shaped structures (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere)"} else { "Figure \\@ref(fig:sphere-proj)"})`) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:sphere-tb-html)', '\\@ref(tab:sphere-tb-pdf)'))` summarizes these functions.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle". In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = 1/3\cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, $X_3 = ru$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional noise dimensions $X_4$ to $X_p$ are generated.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional noise dimensions are added to embed the structure in higher dimensions.

@@

#### Trigonometric

@@

#### Trefoil knots

### 2025-09-19 paper-cardinalR.tex
Updated

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. These shapes help test how well methods preserve symmetry, curvature, and cluster separation. Table \ref{tab:sphere-tb-pdf} summarizes these functions.

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle.'' In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = 1/3\cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), \(X_3 = ru\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional dimensions \(X_4\) to \(X_p\) are generated as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional dimensions \(X_5\) to \(X_p\) are added as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p\).

>>>>

Sphere-shaped structures (Figure \ref{fig:sphere-proj}) are useful for evaluating how dimension reduction and clustering algorithms handle curved, symmetric manifolds in high-dimensional spaces. The package provides functions to generate a variety of spherical forms, including simple circles, uniform spheres, grid-based spheres, and complex arrangements like clustered spheres within a larger sphere. Table \ref{tab:sphere-tb-pdf} summarizes these functions.

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle''. In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = 1/3\cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), \(X_3 = ru\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional noise dimensions \(X_4\) to \(X_p\) are generated.

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional noise dimensions are added to embed the structure in higher dimensions.
